\chapter*{Acknowledgements}
\markboth{Acknowledgements}{}
\label{acknowledgements}

Completing a Ph.D.\ is hard. I am grateful to all those who have helped me reach this point. In particular, I am indebted to my supervisor, Jonathan Gair, for his guidance, suggestions, patience and ability to solve things just in time. I have enjoyed working with him and have learnt much from him. I am grateful to the rest of the Cambridge Gravity Group, in particular, I thank Stephen Taylor for his suggestions regarding adaptive MCMCs; Robert Cole for many conversations on transient resonances, his help in constructing the approximate fit in \secref{location}, and assistance with Mathematica; Christopher Moore for straightening out the issue of noise curves; Ulrich Sperhake for information on numerical relativity, and Priscilla Ca\~{n}izares for her advice on many things.

I have benefited much from the academic environment in Cambridge. I am grateful to Sverre Aarseth for many conversations on stellar dynamics and $N$-body calculations; to Christopher Tout and John Eldridge for discussions pertaining to stellar evolution, and to Donald Lynden-Bell for many interesting conversations and for suggesting that I consider M32 as a source of extragalactic bursts. I also thank George Efstathiou for his advice at various points. I am pleased to thank Dave Green for his advice on apodization, colour schemes and \LaTeX, and Felix Nissen, Alex Rea and Gen Zhang for pondering a variety of questions regarding coding, mathematics and physics. Finally, I thank my fellow students at the Institute for discussions over coffee, post-seminar reviews, journal club debates and many questions in between.

Outside of my immediate bubble, there are many with whom I have exchanged emails, met at conferences, or heard talks from that I would like to thank. I am indebted to Michele Vallisneri for advice on the (im)proper use of Fisher matrices; Tal Alexander and Clovis Hopman for correspondence on their Fokker--Planck calculations, and Michael Kramer for kindly giving me the data used in \figref{pulsar} ahead of its publication. I am grateful to Thomas Sotiriou and Leo Stein for discussions on $f(R)$-gravity; to Tanja Hinderer and Jeanandrew Brink for ideas regarding transient resonances, and to Ilya Mandel and Trevor Sidery for interesting conversations on $k$-d trees. I would like to thank Pau Amaro-Seoane for his comments on stellar dynamics. I also thank Ramesh Narayan for giving me an idea of how to explain the no-hair theorem. Finally, I thank Carlos Sopuerta and Leor Barack for their support.

The work of my doctorate has been supported by many others. My thanks go to Mark Hurn for tracking down many obscure references. I am grateful for all the help of the administrative staff both in department and in College for their wonderful efficiency, especially to Si{\^a}n Owen, Margaret Harding and Rebecca Sawalmeh. I thank the tireless efforts of the High Performance Computing Service to keep the Darwin cluster, which is provided by Dell Inc.\ using Strategic Research Infrastructure Funding from the Higher Education Funding Council for England, up and running. I am also grateful to those who have provided financial support: to the Science and Technology Facilities Council for the first three and a half years of my time, and to the Cambridge Philosophical Society for a further three months, and to Churchill College, the gravitational waves group of the Institut de Ci{\`e}ncies de l'Espai, and the Institute of Physics' C.R.\ Barber Trust for additional travel expenses.

Whilst this dissertation is the culmination of four years of work, it also marks the end of over $21\units{yrs}$ of education. It seems just to thank those who got me this far, in particular Messrs Johnson, McNabb, Whitworth, Maclaren and Trevor, and Drs Green, Richer and Sandeman. I am also grateful to my fellow Churchillian physicists for the many hours solving supervision questions, amongst other things.

My parents deserve much praise for all their support. I thank them for all their assistance and, especially, for their studious proof reading.

Finally, I thank my friends for giving me a happy distraction from my research. Thanks go to those who joined me for pub quizzes, board games and Nintendo; to my flat-mates, office-mates and those who gave me baked goods; to those who hosted barbecues, murder mysteries and house parties; to my dance partners, fellow MCR committee members and the staff of \textit{The Vicious Penguin}; to those who shared celebrations of birthdays, weddings and graduations; to those who visited, whom I visited and who accompanied me on adventures; to those with whom I watched films, theatre and bad TV; to those with whom I enjoyed concerts, ballets and musicals; to those who helped me make igloos, jellies and decorations; to those who dined with me in Hall, be it for Sunday lunch, formal or feast, to those who joined me for quiet evenings in the bar, and to those with whom I have corresponded via email, postcard or letter. I thank all those in College, across Cambridge and the world outside who have helped me build such a wonderful collection of anecdotes.

